The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-21 paper/ms.tex
Fix line indentation throughout

\newcommand{\Gus}[1]{\textcolor{red}{#1}}
\newcommand{\MM}[1]{\textcolor{blue}{#1}}   %mm  Mordecai

@@

% affiliations
\newcommand{\cca}{Center for Computational Astrophysics, Flatiron Institute,
162 5th Ave., New York, NY 10010, USA}
\newcommand{\penn}{Department of Physics \& Astronomy, University of
Pennsylvania, 209 South 33rd St., Philadelphia, PA 19104, USA}
\newcommand{\amnh}{Department of Astrophysics, American Museum of Natural
History, Central Park West at 79th St., New York, NY 10024, USA}
\newcommand{\columbia}{Department of Astronomy, Columbia University, 550 West
120th St., New York, NY 10027, USA}
\newcommand{\victoria}{Department of Physics \& Astronomy, University of
Victoria, 3800 Finnerty Rd., Victoria, BC, V8P 4HN, Canada}
\newcommand{\nyuphys}{Center for Cosmology and Particle Physics, Department of
Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\newcommand{\nyucds}{Center for Data Science, New York University, 60 5th
Ave., New York, NY 10011, USA}
\newcommand{\mpia}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17,
69117 Heidelberg, Germany}

@@

% Orbital properties of stars, computed from their six-dimensional phase space measurements and an assumed Galactic potential, are used to understand the structure and evolution of the Galaxy. Stellar actions, computed from orbits, are invoked as invariant and thus quantitative labels of a star’s orbit.

%   Quantitatively, an offset in the vertical position of ∼ 15 pc (∼ 120 pc) for a thin (thick) disk orbit is sufficient to induce a 25\% systematic error in the vertical action Jz. Such an offset can be trivially generated by local midplane variations. In the FIRE simulations these are on the order of ~100pc at the solar circle. For the Milky Way, from the mean vertical velocity variation of the Milky Way disk of ∼ 5–10 km/s with radius, combined with the dust-map density distribution, we estimate that the Milky Way midplane variations are of order ∼ 60–170 pc. Action calculations which assume the global and local midplanes are identical, are likely to include the induced errors we report, depending on the volume considered. Variation in the local standard of rest causes similar issues. The variation of the midplane must be taken into account when performing dynamical analysis across the large regions of the disk accessible to Gaia and future missions.

@@

%  Furthermore, we
% show that the local midplane varies by $\sim100\,\pc$ at the solar circle in
% Milky Way-mass cosmological zoom-in simulations from the FIRE project. From
% observations that the mean vertical velocity of the Milky Way disk varies by
% $\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
% variations are of order $\sim60\textup{--}170\,\pc$, with similar offsets seen
% in recent three-dimensional dust maps. Thus, current state-of-the-art action
% calculations, which assume the global and local midplanes are identical, are
% likely to include a systematic vertical offset depending on the volume
% considered. Variation in the local standard of rest induces similar issues. The
% variation of the midplane must be taken into account when performing dynamical
% analysis across the large regions of the disk accessible to \textit{Gaia} and
% future missions.

@@

\section{
Azimuthal Midplane Variations}
\label{sec:local_fire}

@@

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the two sets of simulations, we determine the local midplane as a function of azimuth at the solar circle
that an observer might measure if they were situated in each of these
galaxies% 
. Starting from the coordinate system
described in the previous section, which is 
aligned so that the
$z$-coordinate is approximately perpendicular to the disk plane at the solar circle, we place our imaginary
observer at $z=0$ and a galactocentric cylindrical radius of $8.2\,\kpc$ and
vary the azimuth between $0<\phi<2\pi$. 
At equally spaced values of 
$\phi$ we then compute the median $z$ for stars within a cylinder of radius $0.5\,\kpc$ and
height $1\,\kpc$ perpendicular to the fiducial disk and centered on it. We choose to use 50 bins in azimuth, sufficiently few that no cylinder shares stars with its neighbors.
We then re-define the new
midplane of the cylinder to be the median $z$, re-select stars, and iterate
until the median $z$ value converges. We find that only $10$ iterations of
this procedure are necessary for convergence. The resulting median $z$ is
taken to be what our observer would measure as the local galactic midplane at
each $\phi$.

>>>>

@@

\newcommand{\cca}{Center for Computational Astrophysics, Flatiron Institute,
162 5th Ave., New York, NY 10010, USA}
\newcommand{\penn}{Department of Physics \& Astronomy, University of
Pennsylvania, 209 South 33rd St., Philadelphia, PA 19104, USA}
\newcommand{\amnh}{Department of Astrophysics, American Museum of Natural
History, Central Park West at 79th St., New York, NY 10024, USA}
\newcommand{\columbia}{Department of Astronomy, Columbia University, 550 West
120th St., New York, NY 10027, USA}
\newcommand{\victoria}{Department of Physics \& Astronomy, University of
Victoria, 3800 Finnerty Rd., Victoria, BC, V8P 4HN, Canada}
\newcommand{\nyuphys}{Center for Cosmology and Particle Physics, Department of
Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\newcommand{\nyucds}{Center for Data Science, New York University, 60 5th
Ave., New York, NY 10011, USA}
\newcommand{\mpia}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17,
69117 Heidelberg, Germany}

@@

@@

@@

\section{Azimuthal Midplane Variations}
\label{sec:local_fire}

@@

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the two sets of simulations, we determine the local midplane as a
function of azimuth at the solar circle that an observer might measure if they
were situated in each of these galaxies. Starting from the coordinate system
described in the previous section, which is aligned so that the $z$-coordinate
is approximately perpendicular to the disk plane at the solar circle, we place
our imaginary observer at $z=0$ and a galactocentric cylindrical radius of
$8.2\,\kpc$ and vary the azimuth between $0<\phi<2\pi$. At equally spaced
values of $\phi$ we then compute the median $z$ for stars within a cylinder of
radius $0.5\,\kpc$ and height $1\,\kpc$ perpendicular to the fiducial disk and
centered on it. We choose to use 50 bins in azimuth, sufficiently few that no
cylinder shares stars with its neighbors. We then re-define the new midplane
of the cylinder to be the median $z$, re-select stars, and iterate until the
median $z$ value converges. We find that only $10$ iterations of this
procedure are necessary for convergence. The resulting median $z$ is taken to
be what our observer would measure as the local galactic midplane at each
$\phi$.
